---
title: "Lab 3"
subtitle: "ECOL 620 - Applications in Landscape Ecology"
author: "George Woolsey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    # code_folding: hide
editor_options: 
  chunk_output_type: console
knit: (function(inputFile, encoding){ 
    out_dir <- '../';
    rmarkdown::render(inputFile, encoding = encoding, output_file=file.path(dirname(inputFile), out_dir, 'index.html'));
    file.copy(from = "..//index.html", to = '..///data//lab3_george_woolsey.html', overwrite = TRUE)
  })
  
---

# Setup

```{r, include=FALSE, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results='hide'
  , fig.width = 10
  , fig.height = 7
)
```

```{r, eval=T}
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(ggrepel)
library(cowplot)
library(kableExtra)
# spatial analysis
library(USAboundaries)
library(sf)
library(terra)
library(spatstat) #for point pattern analyses
# set seed
set.seed(11)
```

# Rules

1. Using the provided R Markdown template, answer all questions and show your R code where necessary. Note, some questions are just plain text written answers.
2. Complete your assignment using the R markdown file and submit individual assignments to Canvas. Knit your script and submit an .html file on Canvas. Please use  the following naming convention: lab3_firstname_lastname.html (ex. lab3_kyle_horton.html).  Note, we will not grade labs in any other format.

# Question 1

Describe the biological use of Ripley’s K-function and L-function. Differentiate the differences between the two functions. (4 pts)

<span style="color: teal;">
Ripley's K-function and L-function are used to quantify the aggregation of points on a landscape within a circular radius from a given point. The resulting aggregation is then compared to a null model to determine if there is a significant difference between the observed pattern of points and a null model. The Complete Spatial Randomness (CSR) model is a commonly used null model. The CSR assumes that points are distributed over a study area in a completely random fashion where the probability of a point occurring at any given location is equal across the entire study area (i.e., the points come from a uniform distribution). If Ripley's K-value or the L-value are significantly different from the CSR model, then there is evidence that the points are not distributed randomly on a landscape.
</span>

<span style="color: teal;">
Ripley's K-function quantifies the degree of spatial aggregation of points within a circular radius $r$ and then makes a comparison to what we would expect to see given a completely random process (i.e., CSR). As $r$ increases, the expected value of the K function under the CSR will increase exponentially because the area of a circle grows exponentially. The L-function takes the square root of Ripley's K value to make the relationship between the resultant L-value $L(r)$ and the radius $r$ linear. The expected value of the L-function under the CSR follows a line with a slope of 1 and an intercept of 0 such that $L(r) = r$.
</span>

# Question 2

Describe the ecological meaning for the range of values produced by the L-function (i.e., L(r)-r) For instance, what does 0 mean, 1, or -1? (2 pts)

<span style="color: teal;">
Using the L-function, the null hypothesis CSR follows a line with a slope of 1 and an intercept of 0 such that $L(r) = r$. Subtracting $r$ from the result produces $L(r) - r = 0$. Under these conditions, a value of $0$ indicates that the spatial arrangement of the points is completely random across the landscape, consistent with the CSR. When $L(r)-r > 0$, there is evidence for spatial aggregation of points, whereas if $L(r)-r < 0$ there is evidence for uniform (regular) distribution of points.
</span>

# Question 3

Why are edge effect correction factors necessary? (4 pts)

<span style="color: teal;">
Edge effects occur in point pattern analysis near plot boundaries and arise due to missing data outside of the study area (e.g., spatial extent of the study). Edge effects due to arbitrary study area boundaries influence measurements and statistics by not considering the true, full range of data. Correction of edge effects is accomplished by weighting points or using buffers, for example. Edge effect correction methods that utilize weights consider the circumference of circular buffer around a given point that falls outside the spatial extent of the data and then weights each point within the radius proportionally (e.g., if 25% of circumference is *inside* the plot, the weight for points inside is $\frac{1}{0.25}=4$).
</span>

# Cactus data

## Question 4

Include a heat map (using ggplot) of the Ordway-Swisher Biological Station cactus data (e.g., Figure 4.4d). Use a smoothing parameter that best matches (Figure 4.4d). Visually, do the data appear randomly distributed? (4 pts)

```{r}
# load data
cactus <- read.csv("../data/cactus.csv")
boundary <- read.csv("../data/cactus_boundaries.csv",header=T)
#create spatstat objects
ppp.window <- spatstat.geom::owin(
  xrange = c(boundary$Xmin, boundary$Xmax)
  , yrange = c(boundary$Ymin, boundary$Ymax)
)
ppp.cactus <- spatstat.geom::ppp(x = cactus$East, y = cactus$North, window = ppp.window)

#summary information
# summary(ppp.cactus) %>% glimpse()

# the average intensity that summary yields is lambda
lambda <- summary(ppp.cactus)$intensity

# calculate cactus density/intensity and convert to df
cactus_density_df <- ppp.cactus %>% 
  stats::density(bw = 10) %>% # bw = the smoothing bandwidth (smoothing kernel)
  as.data.frame()

# plot heat map
ppp.cactus %>% 
  stats::density(2) %>% 
  as.data.frame() %>%
ggplot(data = .) +
  geom_tile(mapping = aes(x = x, y = y, fill = value)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_viridis_c(option = "plasma") +
  coord_fixed(ratio = 1) +
  labs(
    x = "Easting (m)"
    , y = "Northing (m)"
    , fill = "Opuntia humifusa \n density (m\u00b2)"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom"
    , legend.direction = "horizontal"
    , legend.text = element_text(size = 7, angle = 30, hjust = 0.7)
    , legend.title = element_text(size = 7)
    , axis.title = element_text(size = 7)
    , axis.text.y = element_text(size = 7)
    , axis.text.x = element_text(size = 7, angle = 35, hjust = 0.7)
    , panel.border = element_blank()
  )
```


Eastern prickly pear (*Opuntia humifusa*) at the Ordway-Swisher Biological Station

## Question 5

What conclusion do you draw from the quadrat.test on ppp.cactus with a 4 by 4 quadrat of ? What are the shortcomings of this approach? (4 pts)

```{r}

```

## Question 6

When implementing the isotropic edge effect correction for the L-function for Ordway-Swisher Biological Station cactus data (Line 119), where do you see the greatest observed difference from the uncorrected L-function occur? Include both plots (uncorrected and corrected). Plot with ggplot. (5 pts)

```{r}
#ggplot()+
#  geom_line(data=Lnone, aes(x=r, y=theo-r), colour="red")+
#  geom_line(data=Lnone, aes(x=r, y=un-r), colour="black")+
#  labs(x="r", y="L(r)-r")+
#  theme_classic()+
#  theme(text = element_text(size=15))
```


## Question 7

Using a point-wise envelope (𝛼 level of 0.05, isotropic correction), at what scales(s) do you reject the null hypothesis? In these regions, are the data clumped, randomly distributed, or uniform? No plot needed. (5 pts).

```{r}

```


# Ponderosa Pine data

## Question 8

Examining the Ponderosa pine data included within the spatstat package, how would you characterize the distribution of trees? Does your interpretation change across scale(s)? Use the envelope function to justify your conclusion. Include the point-wise confidence envelope plot using an 𝛼 level of 0.01 and isotropic correction. Include the plot (produced in ggplot) of the tree locations and of the L-function with envelope (10 pts). (See example on Canvas)

```{r}
#ponderosa
#plot(ponderosa)
#summary(ponderosa)
```


# Simulated data

## Question 9

Examine the spatial dispersion of three simulated point distributions (Poisson, uniform, and clustered). Include the point-wise confidence envelope plots at 𝛼 level of 0.01 for the L-function with isotropic correction (using ggplot). In which distribution(s) do you see complete spatial randomness (CSR)? Include a 3 by 2 plot of the point locations and L-function with envelope (See example of Canvas) (10 pts).

```{r}
#set.seed(42)
#poisson distribution
#pois_dis <- rpoispp(100)

#ggplot()+
#  geom_point(data=as.data.frame(pois_dis), aes(x=x, y=y), colour="darkgreen")+
#  labs(y = "northing (no units)", x = "easting (no units)")+
#  theme_bw()+
#  theme(text = element_text(size=15))+
#  coord_equal()+
#  ggtitle("Poisson Distribution")+
#  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
#set.seed(1)
#regular distribution
#regu_dis <- rSSI(0.09, 70)

#set.seed(21)
#clustered distribution
#clust_dist <- rMatClust(30, 0.05, 4)
```


<span style="color: teal;">
asdfsadfasdf
</span>


```{r, include=FALSE, eval=FALSE}
see p. 107 of [Fletcher and Fortin (2018)]

complete spatial randomness (CSR, i.e. a homogenous Poisson process) is a commonly used null model.
under CSR, it is assumed that..lambda is proportional to the area sampled.

random var from poisson distribution with a mean and var of lambda is the number of points, then draw random number of points from a uniform distribution

`spatstat.geom::owin` = Creates an object of class "owin" representing an observation window in the two-dimensional plane

`spatstat.geom::ppp` = Creates an object of class "ppp" representing a point pattern dataset in the two-dimensional plane.

Ripley's K = 
r = radius surrounding points of interest (scale), as we go further and further from point are we getting something random or not. as r gets bigger the K(r) will start to get bigger exponentially because the area of a circle grows exponentially. r is the radius of a circle. are we deviating from the expectation of what lambda should be given the radius defined? point density of data is either higher or lower than we would expect (if black line falls outside of CI) given CSR.

to go to L-function, take the square root of K(r)

red line is what we would expect to see given a completely random process (CSR), if difference between the data (Kr or Lr) and redline, then reject the null hypothesis of CSR




We start with describing the Ripley’s K function (Ripley 1976) because it is probably the most common second-order statistic for point patterns used in ecology. Ripley’s K calculates the degree of spatial aggregation of points within a circle of radius r and contrasts the observed pattern to that expected under CSR (Fig. 4.3a). Ripley’s K is defined as:

$$
K(r) = \frac{E}{\lambda}
$$

where $\lambda = \frac{n}{A}$, and $E =$ # points within radius $r$ of an arbitrarily chosen point (not including that point). Thus, if $K(r)$ is larger than expected for a radius $r$, the observed number of events shows evidence of clustering, whereas if $K(r)$ is smaller than expected for a radius $r$ there is evidence for regularity. Given that the area of a circle is $\pi \cdot r^2$ , the expected number of points in a circle of radius $r$ under CSR is:
```

