---
title: "Lab 3 - Spatial Dispersion and Point Data"
author: "ECOL 620 - Applications in Landscape Ecology"
date: "Spring 2023"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This lab assignment highlights the use of Ripley’s K-function and L-function for the detection of spatial dispersion in point data. We will us the spatstat package. Download Laboratory #3 materials and store files locally on your computer.

Read “Fletcher_2018_CH4” and open “lab3_practice.RMD” in R. The reading materials are particularly dense. You are not expected to follow all of the mathematical formulas – the emphasis is on the ecological interpretation and application. 


```{r}
#load packages
library(spatstat)          #for point pattern analyses; version 1.56-1 used
library(terra)            #for raster covariate data; version 2.6-7 used
library(tidyverse)
```


```{r}
#import the data from directory
cactus <- read.csv("/Users/kylehorton/Desktop/ECOL_620/Course_materials/Week3/Lab3/data_for_lab3/cactus.csv")
boundary <- read.csv("/Users/kylehorton/Desktop/ECOL_620/Course_materials/Week3/Lab3/data_for_lab3/cactus_boundaries.csv",header=T)
```


```{r}
#create spatstat objects
ppp.window <- owin(xrange=c(boundary$Xmin, boundary$Xmax),
                 yrange=c(boundary$Ymin, boundary$Ymax))
ppp.cactus <- ppp(cactus$East, cactus$North, window=ppp.window)
```


```{r}
#plot raw data
ggplot() +
  geom_point(data=cactus, aes(x=East, y=North, size=Area), colour="blue") +
  coord_fixed(ratio = 1)+
  labs(x = "Longitude (m)", y = "Latitude (m)", size=expression(paste("Area (", m^2,")", sep="")))+
  theme_bw()+
  theme(text = element_text(size=15))
```


```{r}
# data for question #4
#summary information
summary(ppp.cactus)
summary(ppp.cactus)$intensity
#the Average intensity that summary yields is λ

#convert to a dataframe
den_cat=as.data.frame(density(ppp.cactus,10))
```


```{r}
#you can plot this natively in R (with the line directly below). I have provided ggplot code to replicate the plot though. 
plot(density(ppp.cactus, 10))

#range of density values
range(round(den_cat$value,3))
```


```{r}
#heat map
ggplot() +
  geom_tile(data=den_cat, aes(x=x, y=y, fill = value)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))+ 
  scale_fill_viridis_c(option = "plasma")+
  coord_fixed(ratio = 1)+
  theme_bw()+
  labs(x = "Easting (m)", y = "Northing (m)", 
       fill= expression(paste("Opuntia \nhumifusa density (", m^2,")", sep="")))+
  theme(legend.title.align=0.5) +
  theme(text = element_text(size=13))
```


```{r}
#contour plot
ggplot() +
  geom_contour(data=den_cat, aes(x=x, y=y, z = value))+
  coord_fixed(ratio = 1)+
  labs(x = "Easting (m)", y = "Northing (m)")+
  theme_bw()+
  theme(text = element_text(size=15))
```


##################################################################

```{r}
#quadrat counts
Q <- quadratcount(ppp.cactus, nx = 4, ny = 4) #counts in 12.5 x 12.5m quadrats
```


```{r}
#plot
plot(ppp.cactus, cex = 2)
plot(Q, add = TRUE, cex = 1)
```


```{r}
#chi-sq test for complete spatial randomness, CSR
quadrat.test(ppp.cactus, nx = 4, ny = 4, method="Chisq", alternative="two.sided")
```


```{r}
set.seed(11)
#Simulate Poisson Point Pattern
pois_dis <- rpoispp(100)
plot(pois_dis)
```


```{r}
quadrat.test(pois_dis, nx = 4, ny = 4, method="Chisq")
```

##############################################
#4.3.4 Univariate point patterns
##############################################

#-----------------------#
#Ripley's K-function & L-function:
#-----------------------#

```{r}
Knone <- Kest(ppp.cactus, correction="none")
#plot K
plot(Knone, legend=T)
```


```{r}
#plot L with 1:1 expectation
Lnone <- Lest(ppp.cactus, correction="none")
plot(Lnone, legend=T)
```


```{r}
#L and K are related by sqrt(Knone$theo/pi)
plot(sqrt(Knone$theo/pi), Lnone$theo) #this is a perfect 1 to 1 line
plot(sqrt(Knone$theo/pi), Lnone$r) #this is a perfect 1 to 1 line
```


```{r}
#plot L with 0 expectation
plot(Lnone, . - r~r, legend=T)
```

```{r}
#isotropic edge correction
Liso <- Lest(ppp.cactus, correction="isotropic")
plot(Liso, . - r~r, legend=T)
```

```{r}
#Monte Carlo simulations to calculate a global and pointwise confidence envelope under CSR
#nsim relates to the alpha level. See page 118 of Fletcher. e.g., rank of 1 and nsim of 199 equates to an alpha of 0.01
Lcsr <- envelope(ppp.cactus, Lest, nsim=199, rank=1, correction="isotropic", global=F)
Lcsr.g <- envelope(ppp.cactus, Lest, nsim=199, rank=1, correction="isotropic", global=T)
```


```{r}
#plot point-wise envelope
plot(Lcsr, . - r~r, shade=c("hi", "lo"), legend=F)
```

```{r}
#plot global envelope
plot(Lcsr.g, . - r~r, shade=c("hi", "lo"), legend=F)
```



