---
title: "Lab 3"
subtitle: "ECOL 620 - Applications in Landscape Ecology"
author: "First Last"
date: "MM/DD/YYYY"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Rules
1. Using the provided R Markdown template, answer all questions and show your R code where necessary. Note, some questions are just plain text written answers.
2. Complete your assignment using the R markdown file and submit individual assignments to Canvas. Knit your script and submit an .html file on Canvas. Please use  the following naming convention: lab3_firstname_lastname.html (ex. lab3_kyle_horton.html).  Note, we will not grade labs in any other format.

## Question 1
Describe the biological use of Ripley’s K-function and L-function. Differentiate the differences between the two functions. (4 pts)

## Question 2
Describe the ecological meaning for the range of values produced by the L-function (i.e., L(r)-r) For instance, what does 0 mean, 1, or -1? (2 pts)

## Question 3
Why are edge effect correction factors necessary? (4 pts)

# Cactus data
## Question 4
Include a heat map (using ggplot) of the Ordway-Swisher Biological Station cactus data (e.g., Figure 4.4d). Use a smoothing parameter that best matches (Figure 4.4d). Visually, do the data appear randomly distributed? (4 pts)

```{r}

```

## Question 5
What conclusion do you draw from the quadrat.test on ppp.cactus with a 4 by 4 quadrat of ? What are the shortcomings of this approach? (4 pts)

```{r}

```

## Question 6
When implementing the isotropic edge effect correction for the L-function for Ordway-Swisher Biological Station cactus data (Line 119), where do you see the greatest observed difference from the uncorrected L-function occur? Include both plots (uncorrected and corrected). Plot with ggplot. (5 pts)

```{r}
#ggplot()+
#  geom_line(data=Lnone, aes(x=r, y=theo-r), colour="red")+
#  geom_line(data=Lnone, aes(x=r, y=un-r), colour="black")+
#  labs(x="r", y="L(r)-r")+
#  theme_classic()+
#  theme(text = element_text(size=15))
```


## Question 7
Using a point-wise envelope (𝛼 level of 0.05, isotropic correction), at what scales(s) do you reject the null hypothesis? In these regions, are the data clumped, randomly distributed, or uniform? No plot needed. (5 pts).

```{r}

```


# Ponderosa Pine data
## Question 8
Examining the Ponderosa pine data included within the spatstat package, how would you characterize the distribution of trees? Does your interpretation change across scale(s)? Use the envelope function to justify your conclusion. Include the point-wise confidence envelope plot using an 𝛼 level of 0.01 and isotropic correction. Include the plot (produced in ggplot) of the tree locations and of the L-function with envelope (10 pts). (See example on Canvas)

```{r}
#ponderosa
#plot(ponderosa)
#summary(ponderosa)
```


# Simulated data
## Question 9
Examine the spatial dispersion of three simulated point distributions (Poisson, uniform, and clustered). Include the point-wise confidence envelope plots at 𝛼 level of 0.01 for the L-function with isotropic correction (using ggplot). In which distribution(s) do you see complete spatial randomness (CSR)? Include a 3 by 2 plot of the point locations and L-function with envelope (See example of Canvas) (10 pts).

```{r}
#set.seed(42)
#poisson distribution
#pois_dis <- rpoispp(100)

#ggplot()+
#  geom_point(data=as.data.frame(pois_dis), aes(x=x, y=y), colour="darkgreen")+
#  labs(y = "northing (no units)", x = "easting (no units)")+
#  theme_bw()+
#  theme(text = element_text(size=15))+
#  coord_equal()+
#  ggtitle("Poisson Distribution")+
#  theme(plot.title = element_text(hjust = 0.5))
```


```{r}
#set.seed(1)
#regular distribution
#regu_dis <- rSSI(0.09, 70)

#set.seed(21)
#clustered distribution
#clust_dist <- rMatClust(30, 0.05, 4)
```
